---
title: "Practical machine learning course project"
author: "Kolchin"
date: '17/04/2019'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(caret)
library(dplyr)
library(ggplot2)
library(readr)

```

## Data loading
```{r load_datasets}
training <- read.csv("pml-training.csv")
testing <- read.csv("pml-testing.csv")
```
The data contains data from accelerometers placed on the belt, forearm, arm, and dumbell provided by 6 participants doing some kinds of physical activity.

Two datasets with training and testing data were loaded from the course' [description page](https://www.coursera.org/learn/practical-machine-learning/supplement/PvInj/course-project-instructions-read-first).

Both datasets contain `r dim(training)[2]` variables. Training dataset contains `r dim(training)[1]` observations, while testing dataset contains `r dim(testing)[1]` observations.

Additional information on the datasets can be obtained [here](http://groupware.les.inf.puc-rio.br/har#weight_lifting_exercises).
